Density-dependent selection and the limits of relative fitness

### 2018-04-28 manuscript.tex
l

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and density will change when selection acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically-stable populations. In most previous models of density-dependent selection, density-regulating traits are subject to density-dependent selection (e.g. ``$K$-selection’’), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection. We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability. Our model incorporates a density-dependent juvenile ``reproductive excess’’, which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection on each trait separately, even for fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (197 words)

@@

Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$. More specifically, as we show below, $b$ and $c$ selection are density dependent, but $d$ selection is not. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

The strength of $b$ selection declines with increasing density. When types differ in $b$ only ($b$-selection), Eq.~\eqref{eq:master} simplifies to Eq.~\eqref{eq:masterequalc}, and absolute fitness can be written as $W_i=(1+\frac{b_i}{\overline{b}}f(\overline{b},N))/d_i$ where $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is a decreasing function of $N$. Thus, the selection factor $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ declines with increasing density: the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies. Here $b_1=1$, $d_1=2$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.}
\end{figure}

@@

We now turn to the issue of how density responds to selection in our variable-density lottery and previous models of selection in density-regulated populations. In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction). In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions. Consequently, $c$-selection does not cause density to change. In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{bsweep.pdf}
\caption{\label{fig:bsweep} Equilibrium $b$ sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general. Panel (b) shows the density-dependent selection factor $(W_2-W_1)/\overline{W}$ predicted by Eq.~\eqref{eq:master} (solid line) compared to the same selection factor with the density-dependence term $f(\overline{b},N)$ held constant at its initial value (dashed line).}
\end{figure}

@@

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}. However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.

>>>>

Selection is commonly described by assigning constant relative fitness values to genotypes. Yet population density is often regulated by crowding. Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait. When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations. These conditions are met in most previous models of density-dependent selection when selection is strong (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak. Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions. We develop a novel model of density-regulated population growth with three ecologically-intuitive traits: fecundity, mortality, and competitive ability. Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density. We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection. Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models. In such cases our model offers a possible alternative to relative fitness.

\noindent (214 words)

@@

Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$. More specifically, as we show below, $b$- and $c$- selection are density-dependent, but $d$-selection is not. Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}. While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.

The strength of $b$-selection declines with increasing density. When types differ in $b$ only ($b$-selection), Eq.~\eqref{eq:master} simplifies to Eq.~\eqref{eq:masterequalc}, and absolute fitness can be written as $W_i=(1+\frac{b_i}{\overline{b}}f(\overline{b},N))/d_i$ where $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is a decreasing function of $N$. Thus, the selection factor $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ declines with increasing density: the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype variant $1$ with at equal frequencies. Here $b_1=1$, $d_1=2$ and $c_1=1$. For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.}
\end{figure}

@@

We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in previous models of selection in density-regulated populations. In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction). In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions. Consequently, $c$-selection does not cause density to change. In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{bsweep.pdf}
\caption{\label{fig:bsweep} Equilibrium $b$-sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general. Panel (b) shows the density-dependent selection factor $(W_2-W_1)/\overline{W}$ predicted by Eq.~\eqref{eq:master} (solid line) compared to the same selection factor with the density-dependence term $f(\overline{b},N)$ held constant at its initial value (dashed line).}
\end{figure}

@@

Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature. Even if reproductive excesses are ubiquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent. For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent. However, this conclusion is sensititive to the assumptions of well-mixed resource competition models. Spatial localization of consumable resources (e.g. for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}. In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).

Moreover, even in well-mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference. Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection. For instance, sexual selection can be viewed as a form of relative interference competition between genotypes. Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like. Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important. Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}. However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.
